---
title: "07-ai-assisted-analysis"
author: "Professor Shannon Ellis"
date: "2026-10-14"

format:
  html: 
    output-file: 07-ai-assisted-analysis.html
  revealjs:
    output-file: 07-ai-assisted-analysis-slides.html
    slide-number: true
    chalkboard: false
    preview-links: auto
    logo: images/cogs137-logo-hex.png
---

```{r packages, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(emo)

knitr::opts_chunk$set(dpi = 300, echo = TRUE, warning = FALSE, message = FALSE)
```

# AI-Assisted Analysis {background-color="#92A86A"}

## Q&A {.scrollable .smaller}

<!-- Copy this box once per question. Put the question after "Q:" in the title and your answer in the body. -->

::: {.callout-note collapse="true" title="Q: "}
A: 
:::

## Course Announcements

**Due Dates**:

-   ⌨️ **Lab 02** due Fri 10/16 (11:59 PM)
-   📚 **HW 02** due Sun 10/18 (11:59 PM)
-   🔘 Lecture Participation survey "due" after class

## Agenda

1.  Use AI to implement Monday's design choices...and check its work
2.  When AI's analysis runs but is wrong
3.  Prompting with the grammar (of graphics & data manipulation)
4.  AI in COGS 137

## Why a whole day on this?

::: incremental
-   AI is *very* good at syntax (which `theme()` argument? which `forcats` function?)
-   AI is *not* responsible for your analysis. You are.
-   Code that **runs** is not the same as code that **answers your question**
-   The skill that matters: knowing what to ask for...and knowing how to check what you got
:::

. . .

Reminder (course policy): think first, try first; always read & understand the output; cite your sources (including AI).

# Part 1: Implement Monday's design choices {background-color="#92A86A"}

## Where we left off

```{r brexit-data, echo = FALSE}
brexit <- tibble(
  opinion = c(
    rep("Right", 664), rep("Wrong", 787), rep("Don't know", 188)
  ),
  region = c(
    rep("london", 63), rep("rest_of_south", 241), rep("midlands_wales", 145), rep("north", 176), rep("scot", 39),
    rep("london", 110), rep("rest_of_south", 257), rep("midlands_wales", 152), rep("north", 176), rep("scot", 92),
    rep("london", 24), rep("rest_of_south", 49), rep("midlands_wales", 57), rep("north", 48), rep("scot", 10)
  )
)
```

```{r brexit-basic, fig.asp = 0.5, out.width = "70%"}
ggplot(brexit, aes(x = opinion)) +
  geom_bar()
```

::: aside
Same YouGov data as Monday's [notes](06-viz.qmd#the-data-code).
:::

## Monday's design decisions

We decided an effective version of this plot should:

::: incremental
1.  Order categories meaningfully (not alphabetically)
2.  Put long categories on the y-axis
3.  Show the regions (facets), with clean labels
4.  Use meaningful colors (and not rely on red/green)
5.  Use informative labels, and drop redundant ones (legend!)
:::

## 🧠 Write the prompt

[What would you ask an AI tool to get this plot? Write it down first. Then compare with a neighbor.]{style="background-color: #ADD8E6"}

. . .

A vague prompt:

```         
make this brexit plot better
```

. . .

A precise prompt:

```         
I have a tibble `brexit` with two character columns: `opinion`
("Right", "Wrong", "Don't know") and `region` (e.g. "rest_of_south").
Using ggplot2, make a bar chart with opinion on the y-axis, one facet
per region in a single row, facet labels cleaned up, fill mapped to
opinion with colors Wrong = "#ef8a62", Right = "#67a9cf",
Don't know = "gray", no legend, and a title + subtitle citing YouGov
(2-3 Sept 2019). Explain what each line does.
```

## What AI gave back (a realistic example)

::: panel-tabset
### Plot

```{r ref.label = "ai-brexit", echo = FALSE, fig.asp = 0.45, out.width = "95%"}
```

### Code

```{r ai-brexit, fig.show = "hide"}
ggplot(brexit, aes(y = fct_infreq(opinion), fill = opinion)) +
  geom_bar() +
  facet_wrap(~ region, nrow = 1) +
  scale_fill_manual(values = c("#ef8a62", "#67a9cf", "gray")) +
  labs(title = "Most Britons regret voting to leave the EU",
       x = "Count", y = "Opinion") +
  theme_minimal()
```
:::

. . .

[🧠 It runs. Is it right? Check it against each of Monday's decisions.]{style="background-color: #ADD8E6"}

## What's wrong?

::: incremental
-   **Colors are silently wrong.** Unnamed `values` are matched to the levels of `opinion` *alphabetically*: "Don't know" got orange, "Right" got blue, "Wrong" got gray.
-   **Facet labels** are still `rest_of_south`, `midlands_wales`...
-   **Redundant legend**: fill and y both show opinion.
-   **The title overclaims.** Is "most" true? Check it:
:::

. . .

```{r}
brexit |>
  count(opinion) |>
  mutate(prop = n / sum(n))
```

. . .

48% is not "most." The AI wrote a headline the data don't support.

## Check it, then fix it

```{r ai-brexit-fixed, fig.asp = 0.45, out.width = "95%"}
ggplot(brexit, aes(y = opinion, fill = opinion)) +
  geom_bar() +
  facet_wrap(~region, nrow = 1, labeller = label_wrap_gen(width = 12)) +
  guides(fill = "none") +
  labs(title = "Was Britain right/wrong to vote to leave EU?",
       subtitle = "YouGov Survey Results, 2-3 September 2019",
       x = NULL, y = NULL) +
  scale_fill_manual(values = c("Wrong" = "#ef8a62",
                               "Right" = "#67a9cf",
                               "Don't know" = "gray")) +
  theme_minimal(base_size = 16) +
  theme(plot.title.position = "plot",
        panel.grid.major.y = element_blank())
```

::: aside
Facet labels still need cleaning. How would you do it? (Hint: `mutate()` + `case_when()` or `fct_recode()` *before* plotting.)
:::

## A checklist for AI-generated code

::: incremental
1.  **Does it run?** (the lowest bar)
2.  **Did it do what I asked?** Check each request, one by one.
3.  **Did it do anything I *didn't* ask?** (dropped rows, changed order, new colors)
4.  **Are the claims supported?** Titles, subtitles, and text are claims. Verify with a quick `count()` / `summarize()`.
5.  **Can I explain every line?** If not, ask (and then check that too)
:::

# Part 2: When the analysis runs but is wrong {background-color="#92A86A"}

## The question

> Among participants who smoked the high-dose joint, do frequent users have higher blood THC than occasional users in the first 30 minutes after smoking?

```{r cs01-data, echo = FALSE}
source("src/cs01_functions.R")
WB <- read_csv("data/Blood.csv") |> clean_cs01("WB")
```

We'll use the (cleaned) whole blood data from CS01 (`clean_cs01()`, same as the `ggplot2` notes).

## AI's answer

::: panel-tabset
### Plot

```{r ref.label = "ai-thc", echo = FALSE, out.width = "70%"}
```

### Code

```{r ai-thc, fig.show = "hide"}
WB |>
  filter(treatment == "13.4% THC (high dose)", time_from_start <= 30) |>
  group_by(group) |>
  summarize(mean_thc = mean(thc, na.rm = TRUE)) |>
  ggplot(aes(x = group, y = mean_thc, fill = group)) +
  geom_col() +
  labs(title = "THC levels by user group",
       y = "Mean THC (ng/mL)")
```
:::

. . .

[🧠 Think on your own, then chat: what would you check before believing this?]{style="background-color: #ADD8E6"}

Put a [green]{.sticky-green} sticky on the front of your computer when you're done discussing. Put a [pink]{.sticky-pink} if you have a question.

## Issue 1: which samples got in?

`time_from_start <= 30` also includes every sample taken *before* smoking (negative times):

```{r}
WB |>
  filter(treatment == "13.4% THC (high dose)", time_from_start <= 30) |>
  count(timepoint)
```

. . .

Half of the rows the AI averaged are **pre-smoking** baselines. That roughly halves both group means.

## Issue 2: what's the unit of analysis?

How many rows does each participant contribute to AI's summary?

```{r}
WB |>
  filter(treatment == "13.4% THC (high dose)", time_from_start <= 30) |>
  count(id, name = "n_rows") |>
  count(n_rows, name = "n_participants")
```

. . .

Almost everyone appears **twice** (once before and once after smoking), so each "mean" blends two different states of the same person. Our question is about people *after* smoking.

. . .

After fixing the filter, re-run this check: it should be one row per participant.

## Issue 3: what does the plot hide?

::: incremental
-   `na.rm = TRUE` *could* silently drop values. Here it dropped none, but you only know that because you checked (`sum(is.na(thc))`)
-   A bar of means hides the spread...and the sample size
-   Redundant fill + legend (sound familiar?)
:::

## A version we can defend

```{r thc-better, out.width = "70%"}
WB |>
  filter(treatment == "13.4% THC (high dose)",
         time_from_start > 0, time_from_start <= 30) |>
  group_by(id, group) |>
  summarize(thc = mean(thc, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = group, y = thc)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.1, alpha = 0.6) +
  labs(title = "Blood THC, 0-30 min after smoking (high dose)",
       subtitle = "One sample per participant, after smoking",
       x = NULL, y = "THC (ng/mL)") +
  theme_minimal(base_size = 14)
```

. . .

Every fix here was a **decision about the question**. AI couldn't make it for us, because we never told it.

# Part 3: Prompting with the grammar {background-color="#92A86A"}

## Vocabulary is leverage

The words you've learned are exactly the words that get better results:

| Instead of...           | Try...                                                            |
|-------------------------|-------------------------------------------------------------------|
| "make it look better"   | "map `opinion` to fill, remove the legend, use `theme_minimal()`" |
| "split it up by region" | "facet by `region` in one row"                                    |
| "only the early ones"   | "filter to `time_from_start > 0 & time_from_start <= 30`"         |
| "average it"            | "one mean per `id`, then compare groups"                          |

## Prompting tips

::: incremental
-   Describe your data: paste `glimpse()` output (column names + types)
-   State the **question**, not just the code you want
-   Ask for code using functions we've learned in class (it will be easier to check!)
-   Ask it to explain each line...then check the explanation
-   Iterate: small, specific follow-ups beat one giant prompt
:::

# Part 4: AI in COGS 137 {background-color="#92A86A"}

## Norms

::: incremental
-   You should be able to answer "yes": *Can I explain each piece of code and each analysis I'm submitting?*
-   **Cite it** (course policy): note the tool and what you used it for, as a code comment or in your text
-   In your case studies, every decision (which samples, which cutoff, which plot) is yours to explain & defend
-   AI is a great first reviewer: "What might be wrong with this analysis?" (and then you decide)
:::

## Your Turn

::: task
1.  Pick a question about the CS01 whole blood data (from the CS01 intro)
2.  Write a *precise* prompt and get an AI tool to write code for it
3.  Run it. Then go through the checklist:
    -   Did it do what I asked? Anything I didn't ask?
    -   Are the claims supported? What's the unit of analysis?
4.  Fix one thing, by hand
:::

Put a [green]{.sticky-green} sticky on the front of your computer when you're done. Put a [pink]{.sticky-pink} if you want help/have a question.

# Recap {background-color="#92A86A"}

-   Can you describe what to check before trusting AI-generated code or plots?
-   Can you explain why code that runs can still be wrong?
-   Can you write a precise prompt using grammar-of-graphics & `dplyr` vocabulary?
-   Do you know how to cite AI assistance in this course?
