---
title: "16-putting-it-together"
author: "Professor Shannon Ellis"
date: "2026-11-18"

format:
  html: 
    output-file: 16-putting-it-together.html
  revealjs:
    output-file: 16-putting-it-together-slides.html
---

```{r packages, echo = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(tidymodels)
library(emo)

knitr::opts_chunk$set(dpi = 300, echo = TRUE, warning = FALSE, message = FALSE)
```

# Putting the Model Together {background-color="#92A86A"}

## Q&A {.scrollable .smaller}

<!-- Copy this box once per question. Put the question after "Q:" in the title and your answer in the body. -->

::: {.callout-note collapse="true" title="Q: "}
A: 
:::

## Course Announcements

**Due Dates**:

-   ⌨️ **Lab 07** due Fri 11/20 (11:59 PM)
-   💻 **CS02** & 🔘 CS02 team eval survey due Sun 11/22 (11:59 PM)
-   📄 **Final Project rough draft** (qmd + HTML in your final project repo) due Sun 11/22 (11:59 PM)
-   🔘 Lecture Participation survey "due" after class

## Agenda

1.  The whole pipeline, on one slide
2.  One pass through it, start to finish (CS02 data)
3.  Common pitfalls
4.  Mapping it onto your CS02 report

## Why today?

Last time students told us:

> "We did the whole recipe lecture in one lecture...a little bit confusing."

. . .

So today: no new functions. Just the **order** things happen in, and **why**.

# The pipeline {background-color="#92A86A"}

## One slide

| Step | Question it answers | `tidymodels` |
|------|---------------------|--------------|
| 1. Split | What data do I *never* peek at until the end? | `initial_split()`, `training()`, `testing()` |
| 2. Recipe | What are my outcome, predictors, and pre-processing steps? | `recipe()`, `update_role()`, `step_*()` |
| 3. Model | What kind of model, which engine, which mode? | `linear_reg()`, `rand_forest()`, `set_engine()`, `set_mode()` |
| 4. Workflow | Recipe + model, bundled together | `workflow()`, `add_recipe()`, `add_model()` |
| 5. Resample | How well does each candidate model do on data it wasn't fit on? | `vfold_cv()`, `fit_resamples()`, `collect_metrics()` |
| 6. Final fit | How does my chosen model do on the test set (once!)? | `last_fit()`, `collect_metrics()` |
| 7. Interpret | What does this mean for my question? | `collect_predictions()`, `vip::vip()`, a plot + a sentence |

[🧠 Before we code: where have we already done each of these steps?]{style="background-color: #ADD8E6"}

# One pass, start to finish {background-color="#92A86A"}

## 1. Split

```{r}
# in your CS02 repo, the file is at "data/pm25_data.csv"
pm <- read_csv("OCS_data/data/raw/pm25_data.csv") |>
  mutate(across(c(id, fips, zcta), as.factor),
         city = case_when(city == "Not in a city" ~ "Not in a city",
                          city != "Not in a city" ~ "In a city"))

set.seed(1234)
pm_split <- initial_split(pm, prop = 2/3)
train_pm <- training(pm_split)
test_pm  <- testing(pm_split)
```

::: aside
Wrangling that doesn't *learn* anything from the data (like recoding `city`) can happen before the split. Anything that learns from the data (correlations, variances, means) goes in the recipe.
:::

## 2. Recipe

```{r}
lm_rec <- recipe(train_pm) |>
  update_role(everything(), new_role = "predictor") |>
  update_role(value, new_role = "outcome") |>
  update_role(id, new_role = "id variable") |>
  update_role("fips", new_role = "county id") |>
  step_rm(county, zcta) |>
  step_dummy(state, city) |>
  step_nzv(all_predictors()) |>
  step_corr(all_numeric_predictors()) |>
  step_lincomb(all_numeric_predictors())
```

. . .

A recipe is a **plan**. Nothing has been computed yet (no `prep()`/`bake()` needed when we use a workflow).

## 3. Model(s)

We'll compare two candidates:

```{r}
lm_model <- linear_reg() |>
  set_engine("lm") |>
  set_mode("regression")

rf_model <- rand_forest(mtry = 10, min_n = 3) |>
  set_engine("randomForest") |>
  set_mode("regression")
```

. . .

Random forest gets its own recipe (it can use categorical variables directly):

```{r}
rf_rec <- recipe(train_pm) |>
  update_role(everything(), new_role = "predictor") |>
  update_role(value, new_role = "outcome") |>
  update_role(id, new_role = "id variable") |>
  update_role("fips", new_role = "county id") |>
  step_rm(county, zcta) |>
  step_novel(state) |>
  step_string2factor(city) |>
  step_nzv(all_predictors()) |>
  step_corr(all_numeric_predictors())
```

## 4. Workflows

```{r}
lm_wflow <- workflow() |>
  add_recipe(lm_rec) |>
  add_model(lm_model)

rf_wflow <- workflow() |>
  add_recipe(rf_rec) |>
  add_model(rf_model)
```

. . .

[🧠 Who can explain the difference between a recipe, a model, and a workflow...in one sentence each?]{style="background-color: #ADD8E6"}

## 5. Resample to choose a model

```{r}
set.seed(1234)
pm_folds <- vfold_cv(train_pm, v = 4)

lm_cv <- fit_resamples(lm_wflow, resamples = pm_folds)
rf_cv <- fit_resamples(rf_wflow, resamples = pm_folds)

bind_rows(
  collect_metrics(lm_cv) |> mutate(model = "linear regression"),
  collect_metrics(rf_cv) |> mutate(model = "random forest")
) |>
  select(model, .metric, mean, std_err)
```

. . .

We choose using **cross-validation on the training data**, *not* the test set.

## 6. Final fit: touch the test set once

```{r}
rf_final <- last_fit(rf_wflow, split = pm_split)

collect_metrics(rf_final)
```

. . .

`last_fit()` fits on all of the training data, then evaluates on the test data. Once.

## 7. Interpret

```{r pred-plot, out.width = "60%"}
collect_predictions(rf_final) |>
  ggplot(aes(x = value, y = .pred)) +
  geom_point(alpha = 0.5) +
  geom_abline(linetype = "dashed") +
  labs(x = "Observed PM2.5 (µg/m³)", y = "Predicted PM2.5 (µg/m³)",
       title = "Random forest predictions on held-out monitors") +
  theme_minimal(base_size = 14)
```

## 7. Interpret (cont.)

```{r}
test_rmse <- collect_metrics(rf_final) |>
  filter(.metric == "rmse") |>
  pull(.estimate)
```

> With what accuracy can we predict US annual average air pollution concentrations?

. . .

On monitors the model never saw, predictions were off by about **`r round(test_rmse, 2)` µg/m³** (RMSE), on a scale where values range from roughly `r round(min(pm$value), 0)` to `r round(max(pm$value), 0)` µg/m³.

. . .

[🧠 Is that good? Good *enough*? For whom? (That's the discussion section of your report.)]{style="background-color: #ADD8E6"}

# Common pitfalls {background-color="#92A86A"}

## Pitfalls we see every quarter

::: incremental
-   **Pre-processing before splitting** (data leakage): the test set "leaks" into your choices
-   **Choosing a model with the test set**: use cross-validation for that; the test set is for the end
-   **Reporting training performance** as if it were test performance
-   **Forgetting `set.seed()`**: results change every time you render
-   **New levels in the test set** (remember `city` on Friday?): `step_novel()` or rethink the variable
-   **RMSE without units or context**: always interpret it in the units of the outcome
:::

# Your CS02 report {background-color="#92A86A"}

## Where each step goes

| Pipeline step | Report section |
|---------------|----------------|
| Split, recipe | Data (wrangling) & Methods |
| Models, resampling | Analysis |
| Final fit | Results |
| Interpret | Results & Discussion (and limitations!) |

## Your Turn

::: task
With your CS02 group:

1.  Fill in the pipeline table for **your** analysis (including your extension)
2.  Which step does your extension change? (new data? new recipe step? new model?)
3.  Where are you currently stuck?
:::

Put a [green]{.sticky-green} sticky on the front of your computer when you're done. Put a [pink]{.sticky-pink} if you want help/have a question.

# Recap {background-color="#92A86A"}

-   Can you list the steps of the modeling pipeline, in order, and say why that order matters?
-   Can you explain the difference between a recipe, a model, and a workflow?
-   Do you know which data (training vs. testing) to use when choosing vs. evaluating a model?
-   Can you interpret RMSE in the context of the CS02 question?
